Generalized measurement invariance tests with application to factor analysis

### 2011-11-17 www/paper.Rnw
touch-ups in references
set line-spacing to 1 (for the moment)



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@76 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
some LaTeX touch-ups:
{theta} -> {\theta}

replaced a few undirected double quotes by directed double quotes

\citeyear constructs replaced by \citeA



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@77 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%\spacing{1}

@@

The assumption that parameters are
invariant across observations is a
fundamental tenet of many statistical models.  A specific type of 
parameter invariance, measurement invariance, has implications for the
general design and use of psychometric scales.
This concept is 
particularly important because violations can render the scales
difficult to interpret.  
That is, if a set of scales violates measurement invariance,
then individuals with the same ``amount'' of a latent variable 
may systematically receive 
different scale scores.
This may lead researchers to conclude subgroup
differences on a wide variety of interesting constructs 
when, in reality, the scales impact the magnitude of the 
differences.  Further, it can be inappropriate to incorporate scales
violating measurement invariance into structural equation
models, where relationships between latent variables are hypothesized.  Horn
and McArdle \citeyear{HorMca92} concisely summarize the impact of
these issues, stating ``Lack of evidence of measurement 
invariance equivocates conclusions and casts doubt on theory in the
behavioral sciences'' (p.~141).  Borsboom \citeyear{Bor06a} further 
notes that researchers often fail to assess whether measurement
invariance holds.

@@

Instead of choosing a specific $\nu$, a natural idea is to compute $\mathit{LR}(\nu)$
for each possible value in some interval $[\underline{\nu}, \overline{\nu}]$ and
reject if their maximum
\begin{equation} \label{eq:maxlr}
  \max_{\nu \in [\underline{\nu}, \overline{\nu}]} \mathit{LR}(\nu)
\end{equation}
becomes large. Note that this corresponds to maximizing the likelihood w.r.t.\
an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. Andrews \citeyear{And93}
showed that the asymptotic distribution is in fact tractable but nonstandard.
Specifically, the asymptotic distribution of (\ref{eq:maxlr}) is the maximum of a
certain tied-down Bessel process whose specifics also depend on the minimal
and maximal thresholds $\underline{\nu}$ and $\overline{\nu}$, respectively.
See Andrews for the original results and further references, and see 
below for more details on the results' application to measurement invariance.

@@

The test problem of the null hypothesis (\ref{eq:h0}) against the alternatives
(\ref{eq:h1}) and (\ref{eq:h1*}), respectively, has been studied extensively
in the statistics and econometrics literature under the label ``structural change
tests'' (see e.g., \citeNP{BroDur75,And93}) where the focus of interest is the
detection of parameter instabilities of time series models ``along'' time.
Specifically, it has been shown (e.g., \citeNP{Nyb89,Han92,HjoKon02,ZeiHor07})
that cumulative sums of the empirical scores follow specific stochastic processes, allowing
us to use them to generally test measurement invariance. Here, we review
some of the main results from that literature and adapt it to the specific challenges
of factor analysis models. More detailed accounts of the underlying structural
change methods include Hjort and Koning \citeyear{HjoKon02} and
Zeileis and Hornik \citeyear{ZeiHor07}.

@@

Inference can then be based on an extension of the usual central limit
theorem. Under the assumption of independence of individuals (implicit
already in Equation~\ref{eq:loglik}) and under
the usual ML regularity conditions (assuring asymptotic normality
of $\hat {\bm \theta}$), Hjort \& Koning \citeyear{HjoKon02} show
that 
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
In words, there are $k$ cumulative score processes, one for each
model parameter.  This collection of processes follows a
multidimensional Brownian bridge as score contributions accumulate in the 
summation from individual 1 (with lowest value of $V$) to individual $n$
(with highest value of $V$).

@@

For the Cram\'{e}r-von Mises test statistic $\mathit{CvM}$, Nyblom \citeyear{Nyb89}
and Hansen \citeyear{Han92} provide small tables of critical values which
have been extended in the software provided by Zeileis \citeyear{Zei06}.
Critical values for the distribution of the maximum LR/Wald/LM tests
are provided by Hansen \citeyear{Han97}. Note that the distribution depends on
the minimal and maximal thresholds employed in the test.

\subsection{Local Alternatives}
Using results from Hjort and Koning \citeyear{HjoKon02} and Zeileis
and Hornik \citeyear{ZeiHor07}, we can also
capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
under specific alternatives 
of parameter instability. In particular, we can assume that the pattern
of deviation $\bm{theta}(v_i)$ can be described as a constant
parameter plus some non-constant deviation $\bm{g}(i/n)$:
\begin{equation}
    \label{eq:local}
  \bm{\theta}(v_i) = \bm{\theta}_0 + n^{-1/2} \bm{g}(i/n).
\end{equation}
In this case, the scores $\bm{s}(\bm{\theta}_0; x_i)$ from
Equation~\eqref{eq:score} do not have zero expectation but rather
\begin{equation}
    \label{eq:localE}
  E[\bm{s}(\bm{\theta}_0; x_i)] = \bm{0} + n^{-1/2}
  \bm{C}(\bm{\theta}_0) \bm{g}(i/n).
\end{equation}
In this case, the covariance matrix $\bm{C}(\bm{\theta})$ from the expected
outer product of gradients is
\begin{equation}
    \label{eq:localS}
  \bm{C}(\bm{\theta}_0) = E[\bm{s}(\bm{\theta}_0; x_i) \bm{s}(\bm{\theta}; x_i)']    
\end{equation}
which, at $\bm{\theta}_0$, coincides with the expected information matrix.

Under the local alternative described above, Hjort and Koning
\citeyear{HjoKon02} show that the process
$\bm{B}(t, \hat{\bm{\theta}})$ behaves asymptotically like
\begin{equation}
    \label{eq:localB}
    \bm{B}^0(t) + \hat{\bm{I}}^{-1/2} \bm{C}(\hat{\bm{\theta}}) \bm{G}^0(t),
\end{equation}
i.e., a zero-mean Brownian bridge plus a term with non-zero mean driven by
$\bm{G}^0(t) = \bm{G}(t) - t \bm{G}(1)$, where $\bm{G}(t) = \int_0^t
\bm{g}(y) dy$.  Hence, unless the local
alternative has $\bm{g}(t) \equiv \bm{0}$, the empirical process $\bm{B}(t,
\hat{\bm{theta}})$ will 
have a non-zero mean.  Hence, the corresponding tests will have non-trivial
power (asymptotically).  We use these results in the Simulation
section to describe the expected behavior of the Brownian bridge.

@@

These attributes give the proposed tests important advantages over
existing tests, as existing measurement invariance methods cannot:
(1) isolate specific parameters violating measurement invariance, or
(2) test measurement invariance for unknown $\nu$. 
In particular, Millsap \citeyear{Mil05} cites
``locating the invariance violation'' as a major outstanding problem
in the field.

In the next section, we employ the proposed tests to analyze existing
data from Wicherts, Dolan, and Hessen \citeyear{WicDol05}.
Following the example, we consider a
full simulation study involving artificial data.

@@

Wicherts, Dolan, and Hessen \citeyear{WicDol05} utilized confirmatory
factor models to study measurement invariance in a series of
general intelligence scales.  They were interested in the
notion of {\em{stereotype threat}}, whereby stereotypes concerning a
subgroup's ability adversely impact the subgroup's performance on tests of that
ability.  The authors specifically focused on stereotypes concerning
ethnic minorities' performance on tests of general intelligence.  In this section, we used the proposed tests to supplement the authors' original analyses.

@@

To study stereotype threat in a measurement invariance framework,
Study 1 of Wicherts et al.\ \citeyear{WicDol05} involved 295 high school
students completing three intelligence tests in two between-subjects
conditions.  Conditions were defined by
whether or not students received primes about ethnic stereotypes prior to testing.
To study the data, Wicherts et al.\ employed a four-group,
one-factor model with the three intelligence tests as manifest variables.
The groups were defined by ethnicity (majority/minority) and by the
experimental manipulation (received/did not receive stereotype prime).
Results indicated that the intelligence tests lacked measurement
invariance, with the minority group receiving stereotype primes being
particularly different from the other three groups on the most
difficult intelligence test.  Most notably, the factor loading and
intercept for the "minority, stereotype prime" group were smaller, and the unique variance
was larger.

As Wicherts et al. describe, one explanation for this pattern of results involves the idea that ability is positively related to "domain identification" \cite{Ste97}: individuals who are high on ability may be more prone to stereotype threat because they are more "tied in" to the domain of interest and are more aware of stereotypes.  To test this idea, we employ a four-group, confirmatory factor model used by Wicherts et al.\ in tandem with the tests proposed in this paper.  Our $V$ variable is participants' grade point averages (which were unused in the original analyses), which will reflect participants' identification with the academic domain.

\subsection{Method}
Wicherts et al.\ tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step 5b (see pp.\ 703--704 of their paper), which involved
across-group parameter restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.  Most notably, the most difficult intelligence test's factor loading, unique variance, and intercept differed for the "minority, stereotype threat" group, as compared to the other groups.
Factor means were also estimated for three of the groups.  The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA, and CFI statistics.  Wicherts et al.\ also had a Step
6 (the final step), where they examined a model that was simpler than
described above and that provided a slightly better fit to the data.
This model restricted factor means to be equal across a subset of the
groups, however, which was not helpful for our purposes (because we are focusing on a single group).

@@

\subsection{Results}
<<<mz_mod>>=
d <- dget("WichertsEtAl2005_Study1.txt")

@@

\subsection{Discussion}
Application of the proposed measurement invariance tests to the
Wicherts et al.\ \citeyear{WicDol05} data allowed us to study the
domain identification's impact on stereotype threat in a
straightforward manner.  The tests focused on a set of group-specific
parameters within a four-group confirmatory factor model, and they
could be carried out using the results from a single estimated model.
The latter fact is important, as other approaches to the problem
described here may adversely impact model fit and degrees of freedom (e.g., if
GPA is inserted directly into the model) or require multiple models to
be estimated (e.g., for LRTs).  Further, through examination of both
test statistics and cumulative score processes, the tests were easily
interpretable from both a theoretical and applied standpoint.  In the
next section, we study the proposed tests' power via simulation.

@@

<<<constraint_study>>=
tests <- c("maxBB", "meanL2BB", "supLM(0.1)")

@@

Relatedly,
S\'{a}nchez \citeyear{San09} describes a general
method for partitioning/segmenting structural equation models within a 
partial least squares framework.  This method involves direct maximization of
the likelihood ratio (i.e., fitting the model for various subgroups
defined by 
$V$ and choosing the subgroups with the largest likelihood ratio).
Thus, unlike the tests described in this paper, this approach does not
provide a formal significance test with a controlled level of Type~I errors.

>>>>

\spacing{1}

@@

The assumption that parameters are
invariant across observations is a
fundamental tenet of many statistical models.  A specific type of 
parameter invariance, measurement invariance, has implications for the
general design and use of psychometric scales.
This concept is 
particularly important because violations can render the scales
difficult to interpret.  
That is, if a set of scales violates measurement invariance,
then individuals with the same ``amount'' of a latent variable 
may systematically receive 
different scale scores.
This may lead researchers to conclude subgroup
differences on a wide variety of interesting constructs 
when, in reality, the scales impact the magnitude of the 
differences.  Further, it can be inappropriate to incorporate scales
violating measurement invariance into structural equation
models, where relationships between latent variables are hypothesized.
\citeA{HorMca92} concisely summarize the impact of
these issues, stating ``Lack of evidence of measurement 
invariance equivocates conclusions and casts doubt on theory in the
behavioral sciences'' (p.~141).  \citeA{Bor06a} further 
notes that researchers often fail to assess whether measurement
invariance holds.

@@

Instead of choosing a specific $\nu$, a natural idea is to compute $\mathit{LR}(\nu)$
for each possible value in some interval $[\underline{\nu}, \overline{\nu}]$ and
reject if their maximum
\begin{equation} \label{eq:maxlr}
  \max_{\nu \in [\underline{\nu}, \overline{\nu}]} \mathit{LR}(\nu)
\end{equation}
becomes large. Note that this corresponds to maximizing the likelihood w.r.t.\
an additional parameter, namely $\nu$. Hence, the asymptotic distribution of the
maximum $\mathit{LR}$ statistic is not $\chi^2$ anymore. \citeA{And93}
showed that the asymptotic distribution is in fact tractable but nonstandard.
Specifically, the asymptotic distribution of (\ref{eq:maxlr}) is the maximum of a
certain tied-down Bessel process whose specifics also depend on the minimal
and maximal thresholds $\underline{\nu}$ and $\overline{\nu}$, respectively.
See Andrews for the original results and further references, and see 
below for more details on the results' application to measurement invariance.

@@

The test problem of the null hypothesis (\ref{eq:h0}) against the alternatives
(\ref{eq:h1}) and (\ref{eq:h1*}), respectively, has been studied extensively
in the statistics and econometrics literature under the label ``structural change
tests'' (see e.g., \citeNP{BroDur75,And93}) where the focus of interest is the
detection of parameter instabilities of time series models ``along'' time.
Specifically, it has been shown (e.g., \citeNP{Nyb89,Han92,HjoKon02,ZeiHor07})
that cumulative sums of the empirical scores follow specific stochastic processes, allowing
us to use them to generally test measurement invariance. Here, we review
some of the main results from that literature and adapt it to the specific challenges
of factor analysis models. More detailed accounts of the underlying structural
change methods include \citeA{HjoKon02} and \citeA{ZeiHor07}.

@@

Inference can then be based on an extension of the usual central limit
theorem. Under the assumption of independence of individuals (implicit
already in Equation~\ref{eq:loglik}) and under
the usual ML regularity conditions (assuring asymptotic normality
of $\hat {\bm \theta}$), \citeA{HjoKon02} show
that 
\begin{equation} \label{eq:fclt}
  {\bm B}(\cdot; \hat {\bm \theta}) ~\overset{d}{\rightarrow}~ {\bm B}^{0}(\cdot),
\end{equation}
where $\overset{d}{\rightarrow}$ denotes convergence in distribution
and ${\bm B}^{0}(\cdot)$ is a $k$-dimensional Brownian bridge.
In words, there are $k$ cumulative score processes, one for each
model parameter.  This collection of processes follows a
multidimensional Brownian bridge as score contributions accumulate in the 
summation from individual 1 (with lowest value of $V$) to individual $n$
(with highest value of $V$).

@@

For the Cram\'{e}r-von Mises test statistic $\mathit{CvM}$, \citeA{Nyb89}
and \citeA{Han92} provide small tables of critical values which
have been extended in the software provided by \citeA{Zei06}.
Critical values for the distribution of the maximum LR/Wald/LM tests
are provided by \citeA{Han97}. Note that the distribution depends on
the minimal and maximal thresholds employed in the test.

\subsection{Local Alternatives}
Using results from \citeA{HjoKon02} and \citeA{ZeiHor07}, we can also
capture the behaviour of the process $\bm{B}(t, \hat{\bm{\theta}})$
under specific alternatives 
of parameter instability. In particular, we can assume that the pattern
of deviation $\bm{\theta}(v_i)$ can be described as a constant
parameter plus some non-constant deviation $\bm{g}(i/n)$:
\begin{equation}
    \label{eq:local}
  \bm{\theta}(v_i) = \bm{\theta}_0 + n^{-1/2} \bm{g}(i/n).
\end{equation}
In this case, the scores $\bm{s}(\bm{\theta}_0; x_i)$ from
Equation~\eqref{eq:score} do not have zero expectation but rather
\begin{equation}
    \label{eq:localE}
  E[\bm{s}(\bm{\theta}_0; x_i)] = \bm{0} + n^{-1/2}
  \bm{C}(\bm{\theta}_0) \bm{g}(i/n).
\end{equation}
In this case, the covariance matrix $\bm{C}(\bm{\theta})$ from the expected
outer product of gradients is
\begin{equation}
    \label{eq:localS}
  \bm{C}(\bm{\theta}_0) = E[\bm{s}(\bm{\theta}_0; x_i) \bm{s}(\bm{\theta}; x_i)']    
\end{equation}
which, at $\bm{\theta}_0$, coincides with the expected information matrix.

Under the local alternative described above, \citeA{HjoKon02} show that the process
$\bm{B}(t, \hat{\bm{\theta}})$ behaves asymptotically like
\begin{equation}
    \label{eq:localB}
    \bm{B}^0(t) + \hat{\bm{I}}^{-1/2} \bm{C}(\hat{\bm{\theta}}) \bm{G}^0(t),
\end{equation}
i.e., a zero-mean Brownian bridge plus a term with non-zero mean driven by
$\bm{G}^0(t) = \bm{G}(t) - t \bm{G}(1)$, where $\bm{G}(t) = \int_0^t
\bm{g}(y) dy$.  Hence, unless the local
alternative has $\bm{g}(t) \equiv \bm{0}$, the empirical process $\bm{B}(t,
\hat{\bm{\theta}})$ will 
have a non-zero mean.  Hence, the corresponding tests will have non-trivial
power (asymptotically).  We use these results in the Simulation
section to describe the expected behavior of the Brownian bridge.

@@

These attributes give the proposed tests important advantages over
existing tests, as existing measurement invariance methods cannot:
(1) isolate specific parameters violating measurement invariance, or
(2) test measurement invariance for unknown $\nu$. 
In particular, \citeA{Mil05} cites
``locating the invariance violation'' as a major outstanding problem
in the field.

In the next section, we employ the proposed tests to analyze existing
data from \citeA{WicDol05}.
Following the example, we consider a
full simulation study involving artificial data.

@@

\citeA{WicDol05} utilized confirmatory
factor models to study measurement invariance in a series of
general intelligence scales.  They were interested in the
notion of {\em{stereotype threat}}, whereby stereotypes concerning a
subgroup's ability adversely impact the subgroup's performance on tests of that
ability.  The authors specifically focused on stereotypes concerning
ethnic minorities' performance on tests of general intelligence.  In this section, we used the proposed tests to supplement the authors' original analyses.

@@

To study stereotype threat in a measurement invariance framework, Study 1 of
\citeA{WicDol05} involved 295 high school students
completing three intelligence tests in two between-subjects conditions. 
Conditions were defined by whether or not students received primes about ethnic
stereotypes prior to testing. To study the data, Wicherts et al.\ employed a
four-group, one-factor model with the three intelligence tests as manifest
variables. The groups were defined by ethnicity (majority/minority) and by the
experimental manipulation (received/did not receive stereotype prime). Results
indicated that the intelligence tests lacked measurement invariance, with the
minority group receiving stereotype primes being particularly different from the
other three groups on the most difficult intelligence test.  Most notably, the
factor loading and intercept for the ``minority, stereotype prime'' group were
smaller, and the unique variance was larger.

As Wicherts et al. describe, one explanation for this pattern of results involves the idea that ability is positively related to ``domain identification''
\cite{Ste97}: individuals who are high on ability may be more prone to stereotype threat because they are more ``tied in'' to the domain of interest and are
more aware of stereotypes.  To test this idea, we employ a four-group, confirmatory factor model used by Wicherts et al.\ in tandem with the tests proposed
in this paper.  Our $V$ variable is participants' grade point averages (GPA, which were unused in the original analyses), which will reflect participants' identification with the academic domain.

\subsection{Method}
Wicherts et al.\ tested a series of confirmatory factor models for
various types of measurement invariance.  We focus on the model used
in their Step 5b (see pp.\ 703--704 of their paper), which involved
across-group parameter restrictions on the factor loadings, unique
variances, intercepts, and factor variances.  These parameters were
typically restricted to be equal across groups, though a subset of the
parameters were allowed to be group-specific upon examination of
modification indices.  Most notably, the most difficult intelligence test's factor loading, unique variance, and intercept differed for the ``minority,
stereotype threat'' group, as compared to the other groups.
Factor means were also estimated for three of the groups.  The model
provided a reasonable fit to the data, as judged by examination of
$\chi^2$, RMSEA, and CFI statistics.  Wicherts et al.\ also had a Step
6 (the final step), where they examined a model that was simpler than
described above and that provided a slightly better fit to the data.
This model restricted factor means to be equal across a subset of the
groups, however, which was not helpful for our purposes (because we are focusing on a single group).

@@

\subsection{Results}
<<mz_mod>>=
d <- dget("WichertsEtAl2005_Study1.txt")

@@

\subsection{Discussion}
Application of the proposed measurement invariance tests to the
\citeA{WicDol05} data allowed us to study the
domain identification's impact on stereotype threat in a
straightforward manner.  The tests focused on a set of group-specific
parameters within a four-group confirmatory factor model, and they
could be carried out using the results from a single estimated model.
The latter fact is important, as other approaches to the problem
described here may adversely impact model fit and degrees of freedom (e.g., if
GPA is inserted directly into the model) or require multiple models to
be estimated (e.g., for LRTs).  Further, through examination of both
test statistics and cumulative score processes, the tests were easily
interpretable from both a theoretical and applied standpoint.  In the
next section, we study the proposed tests' power via simulation.

@@

<<constraint_study>>=
tests <- c("maxBB", "meanL2BB", "supLM(0.1)")

@@

Relatedly, \citeA{San09} describes a general
method for partitioning/segmenting structural equation models within a 
partial least squares framework.  This method involves direct maximization of
the likelihood ratio (i.e., fitting the model for various subgroups
defined by 
$V$ and choosing the subgroups with the largest likelihood ratio).
Thus, unlike the tests described in this paper, this approach does not
provide a formal significance test with a controlled level of Type~I errors.
